---
title: "SAM_Composite"
author: "Amy Hessl"
date: "4/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ncdf4)
library(raster)
library(lubridate)
library(ggplot2)
library(lattice)
library(rnaturalearth)
library(rnaturalearthdata)
library(rasterVis)
library(viridis)
library(seas)
library(s2dverification)
library(reshape2)
library(dplyr)
```

### Data Download from ECHAM5‐wiso (monthly, 1 degree grid)
https://zenodo.org/record/1249604#.XpnVbFNKhTZ

Data sources, years, extent (could be variables in function call someday)
```{r}
netcdf.file <- "data/d18O_precip_mon_1871_2011.nc"

F_yr <- 1958
L_yr <- 2011
years <- seq(F_yr, L_yr)
ext <- extent(143.5, 148.6, -43.8, -39.5) #tas extent (lon_min, lon_max, lat_min, lat_max)
#ext <- extent(0, 180, -90, 0) #SH extent
```

Get a list of the variables to choose from and confirm that the time 
origin and units are appropriate......
```{r}
nc <- ncdf4::nc_open(netcdf.file)

#select the variable
var.name <- names(nc[['var']])[1] #be sure this is the value you want

summary(nc$dim$lon$vals) #if longitude is 0-360 needs to be rotated
tunits <- ncdf4::ncatt_get(nc, "time", "units")
print(tunits)

ncdf4::nc_close(nc)
```

Rotate because the extent is 0-360 not 0-180 (takes a while)
```{r}
dat <- rotate(brick(netcdf.file, varname= var.name))
```

Crop to spatial extent for faster processessing
```{r}
datC <- crop(dat, ext)
```


### Composite Analysis on SAM
Extract Particular Years/Months from datC (climate raster brick)

datY
```{r}
sel_y <- names(datC)[substr(names(datC), 2, 5) %in% years] 
#use that index to subset the raster brick
datY <- subset(datC, sel_y, value=T)
tail(names(datY), 30) 
```

###Need to create a seasonal mean for SAM to match 
Read in Marshall SAM data and subset to period, extract 10 highest/lowest
Change to 95th percentile or 90th percentile using quantile()
Create a montly seasonal mean for SAM 
Make a factor


Monthly (Needs edit):
Wide -> long
Month as a column
If m > 9, year = year + 1
Factor by month into new years
```{r}
sam <- read.table("data/marshallSamMon.txt")
```

```{r}
sam$year <- as.numeric(rownames(sam))
```


```{r}
sam_long <- melt(sam, id.vars = "year")
sam_ordered <- sam_long[order(sam_long$year), ]
sam_ordered$month <- c(1:12)
```

```{r}
for (x in 1:nrow(sam_ordered)){
if (sam_ordered[x,4] > 9) {
  sam_ordered[x,1] <- (sam_ordered[x,1] + 1)
}
}
```

Create annual values from new years:

```{r}
sam_ANN <- sam_ordered %>% group_by(year) %>% summarise(ANN = mean(value))
sam_ANN <- as.data.frame(sam_ANN)
sam_ANN
```

Extract only the growing season:

```{r}
sam_GRO <- sam_ordered %>% filter(month > 9 | month < 3) %>% group_by(year) %>% summarise(ANN = mean(value))
sam_GRO <- as.data.frame(sam_GRO)
sam_GRO
```

Growing season top quantiles:
```{r}
samY <- sam_GRO[sam_GRO$year %in% years, ]

spN <- samY[samY$ANN < (quantile(samY$ANN, 0.1)),] #bottom 10% growing season SAM
spP <- samY[samY$ANN > (quantile(samY$ANN, 0.9)),] #top 10% growing season SAM

#create a logical index for selecting positive and negative SAM years
spN.i <- years %in% spN$year
spP.i <- years %in% spP$year
```

```{r}
spP.i
```


### Better to look at the growing season, rather than calendar year
First create a factor for growth.years where growth.year begins in Oct previous year and runs through Sept of same year. CHANGE THIS TO MEAN OF DEC-FEB? 
```{r}
st.mon <- "10"  
yr_mo_dy <- substr(names(datY), 2, 11)
d <- as.Date(gsub(".", '/', yr_mo_dy, fixed = T)) 
start <- paste0(min(substr(names(datY), 2, 5)),"-", st.mon, "-15")
grow.years <- mkann(d, as.Date(start)) #an index of the growing season years.  note first and last are incomplete as is last year.
```

Before Oct or after Feb = 0
Not an NA when month(d) = 10,11,12,1,2
```{r}
gsmonths = c(3,4,5,6,7,8,9)
gsmonths.i <- month(d) %in% gsmonths
grow.years[gsmonths.i] <- NA
grow.years
```


use this factor to create means by growth.year
```{r}
datX <- stackApply(datY, indices=grow.years, fun=mean, na.rm=TRUE) 
datB <- dropLayer(datX, "level_NA") #remove the layer of NAs
#note layer names of datB
head(names(datB))
```

```{r}
datB
```


Composite function needs an array
```{r}
datA <- as.array(datB)
dim(datA)
```

```{r}
datA
```


First year is 1957-58 but no data for 1957, so cut it
```{r}
datZ <- datA[, , -1]
```
Make sure it worked:
```{r}
datA[,,c(1:3)]
datZ[,,c(1:3)]
```


Composite
```{r}
samNC <- Composite(datZ, occ=spN.i)
samPC <- Composite(datZ, occ=spP.i)
```

########
Convert the resulting arrays into raster object with same specifications as datY


```{r}
samPCr <- raster(samPC$composite[,,1], xmn=143.4375 , xmx=149.0625 , ymn=-43.7298 , ymx=-39.24469 , crs="+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

```{r}
samNCr <- raster(samNC$composite[,,1], xmn=143.4375 , xmx=149.0625 , ymn=-43.7298 , ymx=-39.24469 , crs="+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```

```{r}
samOut <- stack(samPCr, samNCr)
```

```{r}
coasts <- ne_coastline(scale=50, "sp")
coasts <- crop(coasts, ext)
levelplot(samOut, colorkey=list(title="Pos, Neg Anom", space="bottom"), margin=TRUE, 
          names.attr=names(samOut),     par.settings=rasterTheme(viridis_pal(option = "D")(255))) + 
   layer(sp.polygons(coasts, lwd=1)) 
```


Next step: Mask out composite values that have p-values that are insignificant




NEEDS FIXING:

```{r}
samPC.p <- raster(samPC$pvalue[,,1], xmn=143.4375 , xmx=149.0625 , ymn=-43.7298 , ymx=-39.24469 , crs="+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
samNC.p <- raster(samNC$pvalue[,,1], xmn=143.4375 , xmx=149.0625 , ymn=-43.7298 , ymx=-39.24469 , crs="+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
```


```{r}
samPC.p[samPC.p > 0.05] <- NA #mask out insignificant values
samNC.p[samNC.p > 0.05] <- NA
masked.p <- raster::mask(samPCr, samPC.p)
masked.n <- raster::mask(samNCr, samNC.p)
```


```{r}
samMask <- stack(masked.p, masked.n)
```

```{r}
coasts <- ne_coastline(scale=50, "sp")
coasts <- crop(coasts, ext)
levelplot(samMask, colorkey=list(title="Pos, Neg Anom", space="bottom"), margin=TRUE, 
          names.attr=names(samMask),     par.settings=rasterTheme(viridis_pal(option = "D")(255))) + 
   layer(sp.polygons(coasts, lwd=1)) 
```


Other steps: 
Do for southern hemisphere. Change extent at the top
Rotate will be slow. Stackapply will be slow. Maybe do on lab computer


